---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# mousePreopticR

<!-- badges: start -->
<!-- badges: end -->

The goal of mousePreopticR is to enable more convenient sharing/loading of bioinformatics gene expression datasets for analysis in R, in particular mouse hypothalamic preoptic region cell data from [GEO](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE113576) and [DRYAD](https://datadryad.org/stash/dataset/doi:10.5061/dryad.8t8s248).

## Installation

You can install the development version of mousePreopticR from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("wjonasreger/mousePreopticR")
```

## Usage

This is a basic example which shows you how to solve a common problem:

```{r example}
library(mousePreopticR)
## basic example code
```

