---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# mousePreopticR

<!-- badges: start -->
<!-- badges: end -->

The goal of `mousePreopticR` is to enable more convenient sharing/loading of bioinformatics gene expression datasets for analysis in R, in particular mouse hypothalamic preoptic region cell data from experiments by Moffit et al. The data can be found at [GEO](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE113576) and [DRYAD](https://datadryad.org/stash/dataset/doi:10.5061/dryad.8t8s248).

## Installation

You can install the development version of `mousePreopticR` from [GitHub](https://github.com/) with:

```{r github_install, message=FALSE, warning=FALSE, eval=FALSE}
# install.packages("devtools")
devtools::install_github("wjonasreger/mousePreopticR")
```

Installing the package using `devtools::install_github` may likely have issues due to memory size of package data. Alternatively, you can download `mousePreopticR` [here](https://github.com/wjonasreger/mousePreopticR/archive/refs/heads/master.zip). 

You will need to unzip the package folder, remove _"-master"_ from the package folder name, and relocate it to your desired directory, then install it with the following code using _your own path to the package_.

```{r local_install, message=FALSE}
install.packages("/Users/wjonasreger/Documents/mousePreopticR", repos = NULL, type="source")
```

## Usage

This is a basic example which shows you how to setup the package datasets for analysis:

***NOTE:** The package datasets are split components of original large datasets due to GitHub size limits. So, the documented datasets (i.e., `GEO_barcodes`, `GEO_genes`, `GEO_matrix`, `DRYAD_data`) need to be composed once after each install/update of the package. Then package datasets can be loaded anytime after the initial composition step. You may need to restart your R session after composition is completed.*

```{r setup_usage, message=FALSE}
## import packages
library(Matrix)
library(mousePreopticR)
library(monocle3)


## compose datasets (only need to run this once after installation)
# duration: ~3 minutes
composeData()


## load bio datasets
#    - load all datasets: loadData(datasets = "all")
#    - load only bio datasets (Default): loadData(datasets = "bio")
#    - load only example datasets: loadData(datasets = "example")
loadData()


## viewing data
# to learn more about a dataset, use ?data_name (e.g., ?GEO_matrix)
?GEO_matrix


## bioinformatics data structure
cds = new_cell_data_set(expression_data = GEO_matrix,
                        cell_metadata = GEO_barcodes,
                        gene_metadata = GEO_genes)
cds
```

Ok, that's all! You can now play _[scientist](https://youtu.be/RB-RcX5DS5A)_ and do fancy bioinformatics analysis with the `mousePreopticR` datasets, or even use it's convenience functions to manage other datasets as you wish... but that's probably _a lot_ of work. There are more exquisite ways to spend your time such as getting coffee with your friend 😄☕ ☕😄

...You're still here? Ok [watch this](https://youtu.be/xvFZjo5PgG0) for a cackle ;)
