---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# mousePreopticR

<!-- badges: start -->
<!-- badges: end -->

The goal of mousePreopticR is to enable more convenient sharing/loading of bioinformatics gene expression datasets for analysis in R, in particular mouse hypothalamic preoptic region cell data from [GEO](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE113576) and [DRYAD](https://datadryad.org/stash/dataset/doi:10.5061/dryad.8t8s248).

## Installation

You can install the development version of mousePreopticR from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("wjonasreger/mousePreopticR")
```

## Usage

This is a basic example which shows you how to setup the package datasets for analysis:

***NOTE:* The package datasets are split components of original large datasets due to GitHub size limits. So, the documented datasets (i.e., `ex_data`, `ex_matrix`) need to be composed once after each install/update of the package. Then package datasets can be loaded anytime after the initial composition step. You may need to restart your R session after composition is completed.**

```{r example}
# load dependency packages
library(Matrix)
# load mousePreopticR package
library(mousePreopticR)

# compose datasets
composeData()

# load all datasets
#    - load all datasets (Default): loadData(datasets = "all")
#    - load only example datasets: loadData(datasets = "example")
#    - load only bio datasets: loadData(datasets = "bio")
loadData()

# view data (e.g., ex_matrix)
ex_matrix
```



